Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-26 demystify.tex
updates

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. Also, talk about things that adcom members might not know about, e.g., your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Finally, this is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part at the end of the statement, so do not skip it.

@@

\paragraph{English Test} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which is required for assistantships, e.g., TA~\autoref{sec:funding}.

@@

After you submitted your applications, the waiting game begins! For many students, this is an excruciating time. This section provides some information and tips to help you get through this time.

% but you should try to relax and focus on your study and work.  You can also use this time to improve your profile, e.g., by working on research projects, publishing papers, or improving your connections for better LoR writers.

\paragraph{Interviews} Just like a job application, you can get interviews from some schools. The most common case is that a professor is interested in working with you and want to chat to make decision (e.g., to offer RA).
The professor might include one of their current students as part of the interview.

In some case, the interview is done by several professors, e.g., to see if a student fits to their group or to recruit a very strong student to their program. The interview can also be done by adcom members to see if the student fit their program.

@@

\paragraph{Not getting interviews} While it is generally good to get interview, do not feel so despair if you do not get any.  Many programs do not have the tradition of interviewing applicants. For example, at GMU, most of our admitted students with TA do not go through any interviews.

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty setup interviews based on their (very busy) schedule. So do not be surprised if you get an interview invitation in the last minute (also don't get surprised if your interview is on a Uber ride!). And again, some schools do not do interviews at all.

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their wait list).  However, in general, you should hear back from most schools by around March.

@@

\paragraph{Updating your profile} In general, you should not send emails to update your profile.  However, if you have new publications or other achievements, you can send an email asking to update your profile.

\section{Preparing and Tracking Applications}

\paragraph{Shared Spreadsheet} Many students use a (shared) spreadsheet, e.g., Google Sheets/Docs, to help them and their letter writers to keep track of their applications. Here are some information to put on the sheet.
\begin{itemize}
  \item \textbf{Your info}: Full name, email, phone, link to website/CV.  This is helpful for the writers just in case they want to quickly get some information about you.

  \item \textbf{Applications details}: University, Dept, Application System URL, Submission Deadlines, Application Status (e.g., submitted,  rejected, wait-listed, accepted)

  \item \textbf{LoRs}: Writer 1, Writer 2, Writer 3.  If this is a shared document, you might want to omit names and just use writer X. Under each is their status, e.g., sent/not yet/reminder needed, etc.

\end{itemize}

You can then share this sheet with your reference writers and remind them to submit LoRs on your behalf (see \autoref{sec:lor}). Also, you should periodically update your writers with your status; a simple note such as \emph{``I am getting interviewed by prof. X at school Y. Do you have any advice?''} can get you a lot of help.

@@

@@

You can also consider applying to MS programs, which are typically easier to get in (but you need to pay).  If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.  This is a common path for many students.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typicaly no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have higher "activity" or some other hidden fee, or you may be required to pay fees for each credit hour.

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In many cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\section{Buying Computer Equipments}\label{sec:buying-equipment} Many students, including myself, get very excited about getting to buy new computer equipments and electronics when they start their PhD. However, you should check with your adviser first.  Many profs. have funding to buy you a computer and other equipments (e.g., keyboards, headphones, tablets, etc).  Also, do not worry that this is a "waste" of your adviser's money.  This is part of the funding that your adviser gets to support you and your research.  So use it to get what you need to do your research (and it's also OK to use them for personal purposes!).

>>>>

There are various guides on writing SOP, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. Also, talk about things that adcom members might not know about, e.g., your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Finally, this is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part, so do not skip it.

@@

\paragraph{English Test} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which is required for TA (\autoref{sec:funding}).

@@

After you submitted your applications, the waiting game begins! For many students, this is a very stressful time. This section provides some information and tips to help you get through this time.

\paragraph{Interviews} Just like a job application, you might get interviews from schools. The most common case is that a professor is interested in working with you and wants to chat to make decision (e.g., to offer RA). In some cases, the interview is done by several professors, e.g., to see if a student fits to their group or to recruit a very strong student to their program. The interview can also be done by adcom members to see if the student fit their program.
%The professor might include one of their current students as part of the interview.

@@

\paragraph{Not getting interviews} While it is generally good to get interview, not getting an interview does not mean you're out.  Many programs do not have the tradition of interviewing applicants. For example, at GMU, most of our admitted students with TA do not go through any interviews.

However, not getting an interview typically means you do not get directly RA support (\autoref{sec:funding}) by an individual faculty (if they want to pay you to do research for them, then likely they want to talk to you first).  In this case, your application will be reviewed by the adcom and they will make a decision based on your application.

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty setup interviews based on their (very busy) schedule. Do not be surprised if you get an interview invitation in the last minute. Again, some schools do not do interviews at all.

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their waitlist, or they might not send rejection letters).  However, in general, you should hear back from most schools by around March.

@@

\paragraph{Updating your profile} In general, you should not send emails to update your profile.  However, if you have new publications or other big achievements, you can send an email asking to update your profile.

% \section{Preparing and Tracking Applications}

% \paragraph{Shared Spreadsheet} Many students use a (shared) spreadsheet, e.g., Google Sheets/Docs, to help them and their letter writers to keep track of their applications. Here are some information to put on the sheet.
% \begin{itemize}
%   \item \textbf{Your info}: Full name, email, phone, link to website/CV.  This is helpful for the writers just in case they want to quickly get some information about you.

%   \item \textbf{Applications details}: University, Dept, Application System URL, Submission Deadlines, Application Status (e.g., submitted,  rejected, wait-listed, accepted)

%   \item \textbf{LoRs}: Writer 1, Writer 2, Writer 3.  If this is a shared document, you might want to omit names and just use writer X. Under each is their status, e.g., sent/not yet/reminder needed, etc.

% \end{itemize}

% You can then share this sheet with your reference writers and remind them to submit LoRs on your behalf (see \autoref{sec:lor}). Also, you should periodically update your writers with your status; a simple note such as \emph{``I am getting interviewed by prof. X at school Y. Do you have any advice?''} can get you a lot of help.

@@

\paragraph{What else to do?} Make decision (\autoref{sec:accept-postpone-decline})! accept, reject, or defer the offers (if you have multiple options). Ask to meet with potential advisers (e.g., through Open House or separately) and even their students. Ask about computer equipment and software, office space, and other resources; in many case these will be provided for free by your adviser or department (\autoref{sec:buying-equipment}).

@@

You can also consider applying to MS programs, which are typically easier to get in; but you need to pay.  If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.  This is not uncommon for many students.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have higher "activity" or some other hidden fee (e.g., you may need to pay some fees for each credit hour).

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In most cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} Many students, including myself, get very excited about getting to buy new computer equipment and electronics when they start their PhD. However, you should check with your adviser first.  Many profs. have funding to buy you a computer and other equipments (e.g., keyboards, headphones, tablets, etc).  Also, do not worry that this is a "waste" of your adviser's money.  This is part of the funding that your adviser gets to support you and your research.  So use it to get what you need to do your research (and it's also OK to use them for personal purposes!).
